Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis.

### 2024-01-25 paper/paper.tex
updating after giamma revision

  man,floatsintext]{apa6}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{ordinal, likert, simulations, power\newline\indent Word count: X}
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis.},
  pdfauthor={Filippo Gambarota1 \& Gianmarco Altoè1},
  pdflang={en-EN},
  pdfkeywords={ordinal, likert, simulations, power},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

Enter author note here.

The authors made the following contributions. Filippo Gambarota: Conceptualization, Writing - Original Draft Preparation, Writing - Review \& Editing; Gianmarco Altoè: Writing - Review \& Editing, Supervision.

Correspondence concerning this article should be addressed to Filippo Gambarota, Postal address. E-mail: \href{mailto:filippo.gambarota@unipd.it}{\nolinkurl{filippo.gambarota@unipd.it}}

@@

\abstract{%
Abstract here
}

@@

Psychological research make an extensive use of ordinal data. One of the main reason is probably the usage of Likert scales (Likert, 1932). Ordinal data, as defined by Stevens (1946), belongs to a specific type of measurement scale where ordered numbers are assigned to a variable. In contrast to nominal scales, the labels in ordinal scales are ordered. Unlike interval or ratio scales, there is no explicit assumption about the distance between labels. An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement). Answering 4 compared to 2 suggest an higher agreement but we cannot affirm that there is two times the agreement compared to the second answer. Stevens (1946) and Kemp and Grace (2021) suggested that for ordinal variables is appropriate to calculate ranks-based descriptive statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation) and using appropriate inferential tools (Agresti, 2010; e.g., N. Cliff, 1996). This distinction in terms of the appropriateness of certain descriptive statistics is also relevant when modeling data. Treating ordinal data as metric refers to assuming the labels as actual integer numbers thus assuming a fixed and know distance between levels (Liddell \& Kruschke, 2018). More generally, Norman Cliff (2016) suggest that most of the research questions in behavioral sciences can be considered as ordinal (\emph{is the score} \(x\) higher than the score \(y\)?) concerning variables where the most appropriate measurement scale is probably ordinal.

@@

For the tutorial, we assume the reader is already familiar with basic R programming and introductory theory about linear regression. (VEDI SE SCRIVERE MEGLIO O QULCOSA DI PIU). The current tutorial proposed a set of custom R functions (available on Github ) to understand and simulate basic ordinal regression models. More details and extended examples are available on the supplementary materials.

@@

In this section we introduce some notation for the CM that is used through the paper and in the R code (R Core Team, 2023). We proposed a notation as consistent as possible with the literature and the \texttt{ordinal} (Christensen, 2022) R package used in the tutorial. We define \(Y_k\) as the observed ordinal response with \(1, \dots, k\) levels and \(Y^\star\) is the underlying latent variable. The latent variable is segmented using \(k - 1\) thresholds \(\alpha_1, \dots, \alpha_{k - 1}\). Similarly to the generalized linear models framework (e.g., Fox, 2015), we define \(g(x) = \eta\) as the link function that maps probabilities into the linear predictor \(\eta\). To transform back \(\eta\) into probabilities we use the inverse of the link function \(x = g^{-1}(\eta)\). The specific link function define the type of model and require a different R function. For example, when assuming a Gaussian distribution, fitting a model with a \emph{probit} link function entails using the cumulative distribution function \(g(x) = \Phi^{-1}(x) = \eta\) and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined \(x = g^{-1}(\eta) = \Phi(\eta)\). When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\)\footnote{As done by Agresti (2010), when referring to \(P(Y \leq k)\) we are implicitly conditioning on a particular \(x\) value \(P(Y \leq k | x_i)\)}. Equation \eqref{eq:prob-cum-model1} shows the general cumulative model including predictors \(\mathbf{X}\) and regression coefficients \(\boldsymbol{\beta}\). The minus sign in \(\mathbf{X} \boldsymbol{\beta}\) is used to interpret the \(\beta\) as in the standard regression model (Agresti, 2010) where higher \(\beta\) values corresponds to increased probability of responding higher \(k\) categories.

\begin{equation} 
P(Y \leq k) = g^{-1}(\alpha_k - \mathbf{X} \boldsymbol{\beta}), \;\;k = 1, \dots, k - 1
\label{eq:prob-cum-model1}
\end{equation}

@@

{\centering \includegraphics{paper_files/figure-latex/fig-explain-cumulative-1}

@@

\caption{caption here}\label{fig:fig-explain-cumulative}
\end{figure}

@@

\begin{equation} 
Y^\star_i = \mathbf{X}\boldsymbol{\beta} + \epsilon_i
\label{eq:latent-model}
\end{equation}

The crucial part is \(\epsilon_i\) that is the random component of the model coming from a certain probability distribution. For a \emph{probit} model, errors are sampled from a standard Gaussian distribution while for a \emph{logit} model from a standard logistic distribution. Following the notation by Tutz (2022), the observed ordinal value \(Y_i = k\) comes from \(Y^\star_i\) belonging to the interval defined by the thresholds \(Y_i = k \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}\) where \(- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty\).

@@

The cumulative link model implemented in Equations \eqref{eq:prob-cum-model1} and \eqref{eq:latent-model} can be considered the general formulation that requires specifying the link function \(g(\cdot)\) or the errors distribution \(\epsilon_i \sim D(\mu, \sigma^2)\). Among several available functions the \emph{logit} and \emph{probit} models are the most common. The \emph{logit} model assume a \emph{logit} link function and a logistic distribution as latent variable. On the other side, the \emph{probit} model assume a Gaussian distribution.

The two models provide similar results with a different parameters interpretation. In the next sections we will illustrate the differences and simulation strategies. Figure \ref{fig:fig-logit-vs-probit} depicts the two distributions while Table \ref{tab:tab-model-summary} summarise the presented cumulative models with the proposed link function and the corresponding R code.

@@

{\centering \includegraphics{paper_files/figure-latex/fig-logit-vs-probit-1}

@@

\caption{caption here}\label{fig:fig-logit-vs-probit}
\end{figure}

@@

\caption{\label{tab:tab-model-summary}caption here}
\centering
\begin{tabular}[t]{lllll}
\toprule
\multicolumn{1}{c}{} & \multicolumn{2}{c}{Link Function} & \multicolumn{2}{c}{Inverse Link Function} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
Model & Equation & R Code & Equation & R Code\\
\midrule
Cumulative Logit & $\text{logit}(p) = \text{log}(p / (1-p))$ & `qlogis()` & $e^{\text{logit}(p)} / (1 + e^{\text{logit}(p)})$ & `plogis()`\\
Cumulative Probit & $z = \Phi^{-1}(p)$ & `qnorm()` & $\Phi^(z)$ & `pnorm()`\\
\bottomrule
\end{tabular}
\end{table}

@@

We can start by fitting a simple model, highlighting the crucial parameters where the detailed explanation will be expanded in the next sections. Table \ref{tab:tab-dataset-example} contains simulated data from \(n = 100\) participants rating the agreement about a certain item with \(k = 4\) ordered options. The participants are divided into two groups (\(x\)). We can fit a cumulative link model with \texttt{clm()} function and check the model summary.

@@

\caption{\label{tab:tab-dataset-example}caption here}
\centering
\begin{tabular}[t]{llllllll}
\toprule
group & mean & median & sd & Y1 & Y2 & Y3 & Y4\\
\midrule
a & 2.84 & 3 & 1.11 & 8 (p = 0.16) & 11 (p = 0.22) & 12 (p = 0.24) & 19 (p = 0.38)\\
b & 3.02 & 3 & 1.08 & 5 (p = 0.10) & 13 (p = 0.26) & 8 (p = 0.16) & 24 (p = 0.48)\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
\FunctionTok{summary}\NormalTok{(fit)}
\CommentTok{\#\textgreater{} formula: y \textasciitilde{} x}
\CommentTok{\#\textgreater{} data:    dat}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{}  link  threshold nobs logLik  AIC    niter max.grad cond.H }
\CommentTok{\#\textgreater{}  logit flexible  100  {-}128.88 265.77 5(0)  9.09e{-}13 2.2e+01}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Coefficients:}
\CommentTok{\#\textgreater{}    Estimate Std. Error z value Pr(\textgreater{}|z|)}
\CommentTok{\#\textgreater{} xb   0.3153     0.3670   0.859     0.39}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Threshold coefficients:}
\CommentTok{\#\textgreater{}     Estimate Std. Error z value}
\CommentTok{\#\textgreater{} 1|2  {-}1.7503     0.3432  {-}5.100}
\CommentTok{\#\textgreater{} 2|3  {-}0.3791     0.2722  {-}1.393}
\CommentTok{\#\textgreater{} 3|4   0.4381     0.2717   1.612}
\end{Highlighting}
\end{Shaded}

@@

The two main sections of the model summary are the \emph{Coefficients} section reporting the regression coefficients \(\beta\) and the \emph{Threshold} section reporting the \(\alpha\) estimation. Given that \(k = 4\) we have \(k - 1 = 3\) thresholds and a single \(\beta\) associated with the \(x\) effect. As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients where the interpretation depends on the contrast coding (see Schad, Vasishth, Hohenstein, \& Kliegl, 2020). In R the default is the dummy coding where a factor of \(j\) levels is converted into \(j - 1\) dummy variables. By default, the first level of the factor is taken as the reference level and the \(j - 1\) coefficients represent the comparison between the other levels and the reference.

@@

To understand the logit model we need to introduce odds and odds ratio. The odds of a probability \(p\) is defined as \(\frac{p}{1 - p}\) thus the probability of success divided by the probability of failure. The odds takes value ranging from 0 to \(\infty\). For instance, with a probability of \(p = 0.8\), we have odds of \(4\), indicating that there are four successes for each failure. The same as having \(p = 0.2\) and an odds of \(0.25\) means that for each \(0.25\) successes we have a failure or that we have \(4\) failures for each success. When comparing two groups or conditions we can compare the two odds calculating an odds ratio. The odds ratio is the mostly used statistics to compare groups or conditions on a binary outcome. An odds ratio of \(4\) means that the odds of success at the numerator is 4 times higher than the odds of success at the denominator. The Figure \ref{fig:fig-odds-example} shows the relationship between probabilities and odds. The logit transformation is about taking the logarithm of the odds creating a symmetric function ranging from \(-\infty\) to \(\infty\) with \(p = 0.5\) as the midpoint because \(\text{log}(\frac{0.5}{1 - 0.5}) = 0\).

@@

{\centering \includegraphics{paper_files/figure-latex/fig-odds-example-1}

@@

\caption{caption here}\label{fig:fig-odds-example}
\end{figure}

@@

\CommentTok{\# probability of Y = 1 for the group a}
\NormalTok{dat}\SpecialCharTok{$}\NormalTok{yn }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y) }\CommentTok{\# better as number here}
\NormalTok{(p1a }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
\CommentTok{\#\textgreater{} [1] 0.16}

\CommentTok{\# thus P(Y \textgreater{} 1) = 1 {-} p1a}
\FunctionTok{odds}\NormalTok{(p1a)}
\CommentTok{\#\textgreater{} [1] 0.1904762}

\CommentTok{\# odds ratio Y = 1 a vs b}
\NormalTok{(p1b }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
\CommentTok{\#\textgreater{} [1] 0.1}
\FunctionTok{odds}\NormalTok{(p1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(p1b)}
\CommentTok{\#\textgreater{} [1] 1.714286}
\end{Highlighting}
\end{Shaded}

@@

For example the probability of responding \(Y = 1\) in the group ``a'' is 0.16 corresponding to an odds of 0.19. When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y = 1\) we obtain that the group ``a'' has 1.71 times the odds of responding \(Y = 1\) compared to the group ``b''.

For an ordinal variable we have \(k\) levels and we can calculate \(k - 1\) cumulative probabilities (excluding the last associated with \(p = 1\)). Thus we have \(k - 1\) odds ratio or what is called a cumulative odds ratio. Basically we repeat the previous steps but on the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\). The three odds ratios are very similar correspond to the model estimate \(e^\beta_1\), in R \texttt{exp(fit\$beta)} 1.37. In fact, the logit model essentially estimate the cumulative odds ratio.

@@

From the previous example, we noticed that the \(k - 1\) odds ratios are different. However, increasing \(n\) will result in odds ratios that are very similar. This does not happen by chance but from an implicit (until now) assumption made by the CM called \emph{proportional odds} (PO). Following again the taxonomy by Tutz (2022), each of the presented ordinal regression model has a basic version making the PO assumption. There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds}) and partially (\emph{partial proportional odds}) (Peterson \& Harrell, 1990).

>>>>

  man, mask,floatsintext]{apa6}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{ordinal, simulations, power\newline\indent Word count: X}
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis.},
  pdfauthor={Filippo Gambarota1 \& Gianmarco Altoè1},
  pdflang={en-EN},
  pdfkeywords={ordinal, simulations, power},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

The authors made the following contributions. Filippo Gambarota: Conceptualization, Writing - Original Draft Preparation, Writing - Review \& Editing, Methodology, Software; Gianmarco Altoè: Conceptualization, Writing - Review \& Editing, Supervision.

Correspondence concerning this article should be addressed to Filippo Gambarota, Via Venezia 8, 35131, Padova (Italy). E-mail: \href{mailto:filippo.gambarota@unipd.it}{\nolinkurl{filippo.gambarota@unipd.it}}

@@

\abstract{%
Ordinal data are extremely used in psychology. Common examples include Likert items or variables where categories are ordered without assuming a common distance between labels. Despite the prevalence of ordinal data, a majority of published psychological research relies on applying metric models (e.g., standard linear regression) even when the dependent variable is ordinal. This approach leads to several statistical drawbacks. Similar to what happens with binary data (e.g., accuracies in a cognitive experiment), using metric models instead of e.g.~logistic regression is probably caused by several reasons. Firstly, parameter interpretation can be challenging. Ordinal models commonly used a link function (e.g., logit) for model fitting thus transforming the parameters scale. This is even more critical when estimating models with multiple parameters or interactions. Secondly, nowadays journals require power or sensitivity analysis when planning or evaluating a research work. For common statistical models such as t-tests or the analysis of variance, there are tools to easily estimate the statistical power. For ordinal regressions, estimating the statistical power often requires simulating data thus clearly understanding the statistical background and implementing a Monte Carlo simulation. The objective of this tutorial is to present the essential statistical background of ordinal regression, highlighting the research areas that would benefit more by introducing this model. In particular, we present how to interpret the model parameters with categorical and numerical variables also linking them to common standardized effect size measures. Then we propose methods to report the results, especially focusing on graphical representation. Finally, we present how to estimate the statistical power using Monte Carlo simulations with suggestions on setting parameters in a meaningful way. Through the tutorial, we use the R programming language using specific packages for ordinal models and sharing also ad-hoc developed functions supporting each step. For the sake of reproducibility and to empower researchers in analyzing their studies autonomously, all data and scripts used in the article have been made publicly available.
}

@@

Psychological research make an extensive use of ordinal data. One of the main reason is probably the usage of Likert scales (Likert, 1932). Ordinal data, as defined by Stevens (1946), belongs to a specific type of measurement scale where ordered numbers are assigned to a variable. Beyond Likert-like items, in Psychology there are several applications of ordinal scales. For instance, sociodemographic variables like educational levels or socioeconomic status, as well as general ratings such as pain severity, agreement with a statement, or the evaluation of sensory experiences.

In contrast to nominal scales, the labels in ordinal scales are ordered. Unlike interval or ratio scales, there is no explicit assumption about the distance between labels. An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement). Answering 4 compared to 2 suggest an higher agreement but we cannot affirm that there is two times the agreement compared to the second answer. Stevens (1946) and Kemp and Grace (2021) suggested that for ordinal variables is appropriate to calculate ranks-based descriptive statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation) and using appropriate inferential tools (Agresti, 2010; e.g., N. Cliff, 1996). This distinction in terms of the appropriateness of certain descriptive statistics is also relevant when modeling data. Treating ordinal data as metric refers to assuming the labels as actual integer numbers thus assuming a fixed and know distance between levels (Liddell \& Kruschke, 2018). More generally, Norman Cliff (2016) suggest that most of the research questions in behavioral sciences can be considered as ordinal (\emph{is the score} \(x\) higher than the score \(y\)?) concerning variables where the most appropriate measurement scale is probably ordinal.

@@

For the tutorial, we assume that the reader is already familiar with basic R programming and introductory theory about linear regression. Despite the practical and applicative spirit of the tutorial, we still introduced some statistical theory and equations supported by translation into R code. Similar to other tutorials (DeBruine \& Barr, 2021; Gambarota \& Altoè, 2023), we believe this is the best educational approach especially with the involvement of Monte Carlo simulations.

In the first part we introduce the ordinal models explaining the general structure and assumptions. Then we move with model fitting and parameters interpretation. Finally we introduce the simulation approach for common research scenarios and an application to power analysis. We used R {[}Version 4.3.1; R Core Team (2023){]} for the code, figures and tables. Details about R packages an The tutorial is supported by a set of custom R functions available on Github. Other examples and model extensions along with details about used R packages are available on the Github repository (supplementary materials). For the tutorial we used the following packages.

@@

In this section we introduce some notation for the CM that is used through the paper. We proposed a notation as consistent as possible with the literature and the \texttt{ordinal} package (Christensen, 2022). We define \(Y_k\) as the observed ordinal variable with \(k\) levels and \(Y^\star\) is the underlying latent variable. The latent variable is segmented using \(k - 1\) thresholds \(\alpha_1, \dots, \alpha_{k - 1}\). Similarly to the generalized linear models framework (e.g., Fox, 2015), we define \(g(x) = \eta\) as the link function that maps probabilities into the linear predictor \(\eta\). To transform back \(\eta\) into probabilities we use the inverse of the link function \(x = g^{-1}(\eta)\). The specific link function define the type of model and require a different R function. For example, when assuming a Gaussian distribution, fitting a model with a \emph{probit} link function entails using the cumulative distribution function \(g(x) = \Phi^{-1}(x) = \eta\) and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined \(x = g^{-1}(\eta) = \Phi(\eta)\). When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\)\footnote{As done by Agresti (2010), when referring to \(P(Y \leq k)\) we are implicitly conditioning on a particular \(\mathbf{X}\) value \(P(Y \leq k | \mathbf{X}_i)\)}. Equation \eqref{eq:prob-cum-model1} shows the general cumulative model including predictors \(\mathbf{X}\) and regression coefficients \(\boldsymbol{\beta}\). The minus sign in \(\mathbf{X} \boldsymbol{\beta}\) is used to interpret the \(\beta_j\) as in the standard regression models where higher \(\beta\) values corresponds to increased probability of responding higher \(k\) categories (Agresti, 2010).

\begin{equation}
P(Y \leq k) = g^{-1}(\alpha_k - \mathbf{X} \boldsymbol{\beta}) \;\;k = 1, \dots, k - 1
\label{eq:prob-cum-model1}
\end{equation}

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-explain-cumulative-1}

@@

\caption{Relationship between cumulative probabilities and ordinal outcomes. The \(Y^\star\) latent variable is segmented into \(k\) levels using \(k - 1\) thresholds (\(\alpha\)). The distance between thresholds determine the probability of each outcome. The grey thresholds are only used to compute the \(Y = 1\) and \(Y = k\) probabilities.}\label{fig:fig-explain-cumulative}
\end{figure}

@@

\begin{equation}
\mathbf{Y^\star} = \mathbf{X}\boldsymbol{\beta} + \mathbf{\epsilon}
\label{eq:latent-model}
\end{equation}

The crucial part is \(\epsilon\) that is the random component of the model coming from a certain probability distribution. For a \emph{probit} model, errors are sampled from a standard Gaussian distribution while for a \emph{logit} model from a standard logistic distribution. Following the notation by Tutz (2022), the observed ordinal value \(Y_i = k\) comes from \(Y^\star_i\) belonging to the interval defined by the thresholds \(Y_i = k \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}\) where \(- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty\).

@@

The cumulative link model implemented in Equations \eqref{eq:prob-cum-model1} and \eqref{eq:latent-model} can be considered the general formulation that requires specifying the link function \(g(\cdot)\) or the errors distribution \(\epsilon_i \sim D(\mu, \sigma^2)\). Among several available functions the \emph{logit} and \emph{probit} models are the most common. The \emph{logit} model use a \emph{logit} link function and a logistic distribution as latent variable. On the other side, the \emph{probit} model assume a latent Gaussian distribution using the inverse of the cumulative distribution (\(\Phi^{-1}\)) function as link function.

The two models provide similar results with a different parameters interpretation. In the next sections we will illustrate the differences and simulation strategies. Figure \ref{fig:fig-logit-vs-probit} depicts the two distributions while Table L'output è \ref{tab:tab-model-summary-latex} summarise the presented cumulative models with the proposed link function and the corresponding R code.

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-logit-vs-probit-1}

@@

\caption{Difference between \emph{logit} and \emph{probit} models. On the left the probability density function (PDF). On the right the cumulative distribution function (CDF).}\label{fig:fig-logit-vs-probit}
\end{figure}

@@

\caption{\label{tab:tab-model-summary-latex}(ref:tab-model-summary-docx)}
\centering
\begin{tabular}[t]{ccccc}
\toprule
\multicolumn{1}{c}{} & \multicolumn{2}{c}{Link Function $g(.)$} & \multicolumn{2}{c}{Inverse Link Function $g^{-1}(.)$} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5}
\textbf{Model} & \textbf{Equation} & \textbf{R Code} & \textbf{Equation} & \textbf{R Code}\\
\midrule
Cumulative Probit & $z = \Phi^{-1}(p)$ & \texttt{qnorm()} & $p = \Phi(z)$ & \texttt{pnorm()}\\
\bottomrule
\end{tabular}
\end{table}

@@

We can start by fitting a simple model, highlighting the crucial parameters where the detailed explanation will be expanded in the next sections. Table \ref{tab:tab-dataset-example-latex} contains simulated data from \(n = 100\) participants rating the agreement about a certain item with \(k = 4\) ordered options. The participants are divided into two groups (\(x_a\) and \(x_b\)). We can fit a cumulative link model with \texttt{clm()} function and check the model summary.

@@

\caption{\label{tab:tab-dataset-example-latex}(ref:tab-dataset-example-docx)}
\centering
\begin{tabular}[t]{cccccccc}
\toprule
\textbf{Group} & \textbf{Mean} & \textbf{Median} & \textbf{SD} & \textbf{Y1} & \textbf{Y2} & \textbf{Y3} & \textbf{Y4}\\
\midrule
a & 2.16 & 2 & 1.09 & 19 (p = 0.38) & 11 (p = 0.22) & 13 (p = 0.26) & 7 (p = 0.14)\\
b & 3.12 & 4 & 1.1 & 7 (p = 0.14) & 6 (p = 0.12) & 11 (p = 0.22) & 26 (p = 0.52)\\
\bottomrule
\end{tabular}
\end{table}

@@

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
\FunctionTok{summary}\NormalTok{(fit)}
\CommentTok{\#\textgreater{} formula: y \textasciitilde{} x}
\CommentTok{\#\textgreater{} data:    dat}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{}  link  threshold nobs logLik  AIC    niter max.grad cond.H }
\CommentTok{\#\textgreater{}  logit flexible  100  {-}127.04 262.08 4(0)  5.42e{-}07 2.0e+01}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Coefficients:}
\CommentTok{\#\textgreater{}    Estimate Std. Error z value Pr(\textgreater{}|z|)    }
\CommentTok{\#\textgreater{} xb   1.5929     0.3895    4.09 4.32e{-}05 ***}
\CommentTok{\#\textgreater{} {-}{-}{-}}
\CommentTok{\#\textgreater{} Signif. codes:  0 \textquotesingle{}***\textquotesingle{} 0.001 \textquotesingle{}**\textquotesingle{} 0.01 \textquotesingle{}*\textquotesingle{} 0.05 \textquotesingle{}.\textquotesingle{} 0.1 \textquotesingle{} \textquotesingle{} 1}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Threshold coefficients:}
\CommentTok{\#\textgreater{}     Estimate Std. Error z value}
\CommentTok{\#\textgreater{} 1|2  {-}0.4243     0.2745  {-}1.546}
\CommentTok{\#\textgreater{} 2|3   0.4293     0.2711   1.584}
\CommentTok{\#\textgreater{} 3|4   1.5851     0.3165   5.008}
\end{Highlighting}
\end{Shaded}

The two main sections of the model summary are the \emph{Coefficients} section reporting the regression coefficients \(\beta\) and the \emph{Threshold} section reporting the \(\alpha\) estimation. Given that \(k = 4\) we have \(k - 1 = 3\) thresholds and one \(\beta\) associated with the \(x\) effect. As in standard regression models, when \(x\) is a categorical predictor with \(j\) level, we will estimate \(j - 1\) regression coefficients (plus the intercept term) where the interpretation depends on the contrast coding (see Schad, Vasishth, Hohenstein, \& Kliegl, 2020). In R the default is the dummy coding where a factor of \(j\) levels is converted into \(j - 1\) dummy variables. By default, the first level of the factor is taken as the reference level and the \(j - 1\) coefficients represent the comparison between the other levels and the reference.

@@

To understand the logit model we need to introduce odds and odds ratio. The odds of a probability \(p\) is defined as \(p/(1 - p)\) thus the probability of success divided by the probability of failure. The odds takes value ranging from 0 to \(\infty\). For instance, with a probability of \(p = 0.8\), we have odds of \(4\), indicating that there are four successes for each failure. The same as having \(p = 0.2\) and an odds of \(0.25\) means that for each \(0.25\) successes we have a failure or that we have \(4\) failures for each success. When comparing two groups or conditions we can compare the two odds calculating an odds ratio. The odds ratio is the mostly used statistics to compare groups or conditions on a binary outcome. An odds ratio of \(4\) means that the odds of success at the numerator is 4 times higher than the odds of success at the denominator. The Figure \ref{fig:fig-odds-example} shows the relationship between probabilities and odds. The logit transformation is about taking the logarithm of the odds creating a symmetric function ranging from \(-\infty\) to \(\infty\) with \(p = 0.5\) as the midpoint because \(\text{log}(0.5/(1 - 0.5)) = 0\).

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/fig-odds-example-1}

@@

\caption{Relationship between probability (x axis) odds and the log odds. Using log odds ensure a symmetric relationship with a zero midpoint (\(p = 0.5\)).}\label{fig:fig-odds-example}
\end{figure}

@@

\CommentTok{\# probability of Y = 1 for the group a}
\NormalTok{dat}\SpecialCharTok{$}\NormalTok{yn }\OtherTok{\textless{}{-}} \FunctionTok{as.integer}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{y) }\CommentTok{\# better as number here}
\NormalTok{(p1a }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
\CommentTok{\#\textgreater{} [1] 0.38}

\CommentTok{\# thus P(Y \textgreater{} 1) = 1 {-} p1a}
\FunctionTok{odds}\NormalTok{(p1a)}
\CommentTok{\#\textgreater{} [1] 0.6129032}

\CommentTok{\# odds ratio Y = 1 a vs b}
\NormalTok{(p1b }\OtherTok{\textless{}{-}} \FunctionTok{mean}\NormalTok{(dat}\SpecialCharTok{$}\NormalTok{yn[dat}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{] }\SpecialCharTok{==} \DecValTok{1}\NormalTok{))}
\CommentTok{\#\textgreater{} [1] 0.14}
\FunctionTok{odds}\NormalTok{(p1a) }\SpecialCharTok{/} \FunctionTok{odds}\NormalTok{(p1b)}
\CommentTok{\#\textgreater{} [1] 3.764977}
\end{Highlighting}
\end{Shaded}

@@

For example the probability of responding \(Y = 1\) in the group ``a'' is 0.38 corresponding to an odds of 0.61. When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y = 1\) we obtain that the group ``a'' has 3.76 times the odds of responding \(Y = 1\) compared to the group ``b''.

For an ordinal variable we have \(k\) levels and we can calculate \(k - 1\) cumulative probabilities (excluding the last associated with \(P = 1\)). Thus we have \(k - 1\) odds ratio or what is called a cumulative odds ratio. Basically we repeat the previous steps but on the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\). The three odds ratios are very similar correspond to the model estimate \(e^\beta_1\), in R \texttt{exp(fit\$beta)} 4.92. In fact, the logit model essentially estimate the cumulative odds ratio.

@@

From the previous example, we noticed that the \(k - 1\) estimated odds ratios are different. However, increasing \(n\) will result in odds ratios that are very similar. This does not happen by chance but from an implicit (until now) assumption made by the CM called \emph{proportional odds} (PO). Following again the taxonomy by Tutz (2022), each of the presented ordinal regression model has a basic version making the PO assumption. There are more advanced versions of the model relaxing this assumption completely (\emph{non proportional odds}) and partially (\emph{partial proportional odds}, Peterson \& Harrell, 1990).

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{set.seed}\NormalTok{(}\DecValTok{2024}\NormalTok{)}
\CommentTok{\# using same parameters as the previous simulation increasing n}
\NormalTok{n }\OtherTok{\textless{}{-}} \FloatTok{1e5} \CommentTok{\# very large n}
\NormalTok{x }\OtherTok{\textless{}{-}} \FunctionTok{rep}\NormalTok{(}\FunctionTok{c}\NormalTok{(}\StringTok{"a"}\NormalTok{, }\StringTok{"b"}\NormalTok{), }\AttributeTok{each =}\NormalTok{ n)}
\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{x =}\NormalTok{ x)}
\NormalTok{dat }\OtherTok{\textless{}{-}} \FunctionTok{sim\_ord\_latent}\NormalTok{(}\SpecialCharTok{\textasciitilde{}}\NormalTok{x, }\AttributeTok{beta =}\NormalTok{ b1, }\AttributeTok{prob0 =}\NormalTok{ probs, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
